---
title: "Installing and setting up OpenRouteService locally"
author: Jonas Lieth
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing and setting up OpenRouteService locally}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup}
library(ORSRouting)
```

# Why do the routing functions not work off the shelf?

First off, ORSRouting is powered by OpenRouteService (ORS) and as such depends on a running ORS instance. If you did not mount a working ORS instance first, all routing functions will throw an error. An OpenRouteService instance is a server that can be communicated with using a specified REST API. This can either be the [public API](https://api.openrouteservice.org), a local server or any kind of web server that runs ORS. Instances are controlled through the `ors_instance` functions and the ORS setup function family. The resulting `ors_instance` object is automatically stored in a package environment and detected by all routing functions in the package. If you have no interest in setting up a local instance and just want to make quick routing requests, you can grab a free API key from the [official website](https://openrouteservice.org/) and chuck it into the `ORS_TOKEN` environment variable:

```{r public, eval=FALSE}
Sys.setenv(ORS_TOKEN = "API token goes here")

# `server = "api"` is short for https://api.openrouteservice.org/
public <- ors_instance(server = "api")
```

After doing this, all functions should be ready to use. Keep in mind though, that firing requests at the public API comes with lots of rate restrictions. For a simple request involving 100 routes you will wait upwards of two minutes as requests are throttled according to their rate restrictions. If you plan on working with larger datasets, if you're impatient, or if you're handling sensitive data, it is strongly suggested that you set up a local instance. This vignette will cover the usage of `ors_instance` and family to construct and start a customized ORS instance using R.



# Setting up the service backend

In order to set up the ORS backend locally, three requirements must be met:

- Docker (Docker Desktop on Windows)
- An OpenStreetMap extract of the desired region
- The source code of OpenRouteService

Of these, only Docker needs to be installed beforehand. `ORSRouting` can verify the installation and start Docker, but it cannot assist in the installation process. OpenRouteService will sit in a container which is accessible over a localhost address. After installing Docker, a new ORS instance can be initialized. If necessary, this will download the ORS GitHub repository and start the Docker daemon.

```{r init, eval=FALSE}
ors <- ors_instance(dir = "~", version = "6.7.0", verbose = FALSE)
ors
```

The resulting `ors_instance` object is stored in a cache environment for the entire R session. If the `ors` instance needs to be updated (when changing a file outside of R) or renewed (when starting a new R session), you can use the same function:

```{r renew, eval=FALSE}
ors <- ors_instance(ors)
```

The `ors` object holds all information about the instance and its current configuration. `$paths` contains information about all relevant directories and files. `$extract` shows the mounted extract. `$config` contains the parsed configuration file. `$settings` contains the parsed `docker-compose.yml` file. `$status` displays the current status of the Docker container. Each of these elements can be modified using the ORS setup family functions. All of these functions are designed to be pipeable and the entire setup can be reduced to a (admittedly pretty long) one-liner:

```{r all-in-one, eval=FALSE}
ors <- ors %>%
    ors_extract("Rutland") %>%
    ors_config(
      profiles = c("car", "bike-regular", "walking"),
      maximum_snapping_radius = -1,
      maximum_distance = 100000
    ) %>%
    ors_settings(
      name = "ors_stable",
      ports = NA,
      memory = c(0.1, 0.2),
      auto_deletion = FALSE
    ) %>%
    ors_up()
```


This code chuck follows a rather simple workflow:

1. Create an instance
2. Mount an OpenStreetMap extract
3. Adjust the routing configuration
4. Change the Docker container settings
5. Start the service by building the container

You can choose to either mount a local extract or download one by providing a place name. Extract downloads are powered by the `oe_match` and `oe_download` functions from the [osmextract](https://docs.ropensci.org/osmextract/) package and follow their logic.

Adjusting the routing configuration means making changes to the configuration file of OpenRouteService. The config file controls how routes are computed. The most common options are included as arguments for convenience. For example, in the chunk above, the maximum number of of routes that are allowed to be computed at once is increased to 10,000. For more specified changes, you can choose to provide a modified version of the parsed json file (as returned by `jsonlite::toJSON`) or change the file in an editor and update the instance. One central part of the configuration file are routing profiles. A profile is a means of transporation that is used as a basis for routing. Each profile needs its own graphs built and thus increases the time and memory that a container needs to be built. You should therefore limit the profiles you activate in a single instance based on the amount of available RAM. In the code chunk, two profiles are activated: `driving-car` and `cycling-regular`.

Changing the Docker settings means editing the Docker compose file (`docker-compose.yml`) in the ORS directory. Again, the most important settings are included as arguments (project name, ports, memory and graph building). Other relevant options (if there are any) can be provided through a parsed yaml file (as returned by `yaml::as.yaml`) or by modifying the compose file directly and updating the instance. Keep in mind that RAM is an important resource for local ORS instances and the more profiles you build the more RAM you have to allocate. Not allocating enough RAM leads to the container build failing.

The Docker container can be controlled through a purpose-tailored low-level Docker interface that understands the following commands:

- Build and start a container (`ors_up` / `docker-compose up -d`)
- Take a container down (`ors_down` / `docker-compose down`)
- Start a container (`ors_start` / `docker-compose start`)
- Stop a container (`ors_stop` / `docker-compose stop`)


